---
title: "Ecological valuation assessment"
author: "Marie-Catherine Bouquieaux, Willem Boone, Hanneloor Heynderickx"
date: "01/10/2024"
output:
  pdf_document: 
    toc: true
html_document: default
editor_options: 
  markdown: 
    wrap: 60
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, tidy.opts=list(width.cutoff=45),tidy=TRUE)
```

#### Project: MARBEFES

#### Copyright (C) 2024 by VLIZ

**Contact: Marie-Catherine Bouquieaux ([marie.catherine.bouquieaux\@vliz.be](mailto:marie.catherine.bouquieaux@vliz.be){.email})**\
**Vlaams Instituut voor de Zee/Flanders Marine Institute**\
**Jacobsenstraat 1**\
**8400 Oostende, Belgium**

## 1. Install & load required libraries

If you like to knit the Rmarkdown script into a PDF an additional installation of a LaTeX program is required. Option to use tinytex::install_tinytex(). In addition, this Rmarkdown was created and ran using the following version:

-   R: 4.3.2

-   Rstudio: 2023.12.1

-   Computer: Windows 11

-   Package: last version available on the 08 of March 2024, except for ggplot2, for which the version is specified below

*List of packages used*
```{r list_package, message=FALSE, warning=FALSE}
my_packages <- c( "data.table","sf", "raster",   
                 "tidyr","dplyr", "rnaturalearth","rnaturalearthdata", "devtools", "knitr","kableExtra", "ggplot2") 
```

*To run only if some packages mentioned above are not already installed*
```{r install_package}
install.packages(my_packages)
```

*Load packages*
```{r load_package, message=FALSE, warning=FALSE}
lapply(c(my_packages), require, character.only = TRUE)
```

## 2. Input parameters (Settings)

###2.1 Define the working environment

**Input**: Filepath containing the shapefile **and** shx, .dbf and .prj for each Ecological Component.

**Output directory**: Filepath for output files; make sure directory exists.

**list_EC_result**: List of the files with the EC results; please add the extension of the file as well (for example "EVA_EC_bird_layers.shp")

**xmin/xmax**: Minimum and maximum longitude value to be used as limit on the map; must be provided in degrees.

**ymin/ymax**: Minimum and maximum latitude value to be used as limit on the map; must be provided in degrees.

**quantile_classification**: Please indicate TRUE if you prefer to utilize a distribution-based method for classifying your results, where the categories are determined by quantiles (20,40,60,80,100). If FALSE is enter, specific categories will be used and can be changed/defined in point 7. 

**CreateGrid:** create a grid for subzones. TRUE or FALSE parameter. (alternative to using shapefile polygons).

```{r setting1}
## Path settings for input and output data
Input<-"~/Documents/EVA_process/output/"
Output<-"~/Documents/EVA_process/final_EV/"

## Test if output path exists, if not create it 
if (!dir.exists(Output)) {
  dir.create(Output, recursive = TRUE)
}

# List of files: 
list_EC_result_3km <- c("EVA_EC_bird_layers.shp","EVA_EC_fish_layers.shp","EVA_EC_mammal_layers.shp","EVA_EC_phytoplankton_layers.shp","EVA_EC_zooplankton_layers.shp")
list_EC_result_250m <- c("EVA_EC_benthos_layers.shp")

#Max/min latitude and longitude
xmin <- 1.5
xmax <- 3.5
ymin <- 51
ymax <- 55.5 


## Path settings for the coordinates of the BBT
# Complete path and file name containing the BBT information
BBT_coordinates <- "~/Documents/EVA_process/BBT_coordinates/centerlinev2_20kmbuffer.gpkg"
# If a specific layers need to be used, enter name here:
BBT_layer <- ""

# Create grid? 
CreateGrid<-TRUE

#Show result with continous scale or with discrete variables
continous <- TRUE

```

### 2.2 Get surrounding country map information and BBT/grid coordinates

Loading the coordinates of the BBT and compute its
projection in the appropriate format.

```{r bbtcoord, message=FALSE, warning=FALSE, results='hide'}
# You should have specify the path/name of your file above, as well as the layer name is any is present
if (nchar(BBT_layer) > 0){
  polygon_bbt <- st_read(BBT_coordinates, layer = BBT_layer)
} else {
  polygon_bbt <- st_read(BBT_coordinates)
}
# Change projection of the coordinates to ease the grid calculation
Polygon_coordinated_32631 <- st_transform(polygon_bbt, crs = st_crs(32631))

```

**Get surrounding country map information**

```{r world_data}
world_data <- ne_countries(scale = "medium", returnclass = "sf") |>
  sf::st_transform(crs="EPSG:4326") |>
  sf::st_crop(c(xmin= xmin, xmax = xmax, ymin = ymin, ymax = ymax))
```

**Get GRID coordinates**

```{r bbtcoord_map}
Polygon_coordinated_4326 <- Polygon_coordinated_32631 %>% 
  sf::st_transform(crs="EPSG:4326")

if(CreateGrid==FALSE){ #to check
  # If your file contains different layers, specify which one should be used by entering the option : layer = "name"'
  # If there are no different layers present, remove the option layer = "name"
  GRD <- st_read(SubzoneShapeName)
  # Change projection of the coordinates to ease the grid calculation.
  Polygon_coordinated_32631 <- st_transform(polygon_bbt, crs = st_crs(32631))
  # Create a dataframe with both the grid_polygon and grid_id for each subzone.
  grid_polygons_sf <- st_sf(grid_id = seq_along(GRD[Polygon_coordinated_32631$geom]), 
                          geometry = GRD[Polygon_coordinated_32631$geom], crs = 32631)
  
} else if(CreateGrid==TRUE){
  # Create grid 
  GRD <- st_make_grid(Polygon_coordinated_32631$geom, cellsize = 3000, square = FALSE)
  
  # Create a dataframe with both the grid_polygon and grid_id for each subzone.
  grid_polygons_sf <- st_sf(grid_id = seq_along(GRD[Polygon_coordinated_32631$geom]), 
                          geometry = GRD[Polygon_coordinated_32631$geom], crs = 32631)
}

GRID_degree_4326 <- st_transform(grid_polygons_sf,crs="EPSG:4326")

```

## 3. Rescaling of the 250m grid

If you have EC that have been calculated using a 250m grid, please run the following code to resize it to a 3km grid.

### 3.1 Resize the smaller grid

```{r}
if ( any(nchar(list_EC_result_250m) > 0 )){
  # You should have specify the path/name of your file above, as well as the layer name is any is present
  if (nchar(BBT_layer) > 0){
    polygon_bbt <- st_read(BBT_coordinates, layer = BBT_layer)
  } else {
    polygon_bbt <- st_read(BBT_coordinates)
  }
  # Change projection of the coordinates to ease the grid calculation
  Polygon_coordinated_32631 <- st_transform(polygon_bbt, crs = st_crs(32631))
  
  
  # 250m grid
  GRD_250 <- st_make_grid(Polygon_coordinated_32631$geom, cellsize = 250, square = FALSE)
  # Create dataframe with both the grid polygon and grid id for each subzones
  grid_polygons_sf_250 <- st_sf(grid_id = seq_along(GRD_250[Polygon_coordinated_32631$geom]),
                            geometry = GRD_250[Polygon_coordinated_32631$geom], crs = 32631)
  # 3km grid
  GRD_3km <- st_make_grid(Polygon_coordinated_32631$geom, cellsize = 3000, square = FALSE)
  # Create dataframe with both the grid polygon and grid id for each subzones
  grid_polygons_sf_3km <- st_sf(grid_id_3km = seq_along(GRD_3km[Polygon_coordinated_32631$geom]),
                            geometry = GRD_3km[Polygon_coordinated_32631$geom], crs = 32631)
  
  hexagons_250 <- st_transform(grid_polygons_sf_250, crs = "EPSG:4326")
  hexagons_3000 <- st_transform(grid_polygons_sf_3km, crs = "EPSG:4326")
  
  #250m grid associated to one 3km hexagon
  inside_grid <- st_join(st_centroid(hexagons_250),hexagons_3000, join = st_intersects) 
  inside_grid <- st_drop_geometry(inside_grid) 
  
  
  data_EC_rescaled <- data.frame()
  
  for (j in length(list_EC_result_250m)){
    data_to_rescale <- st_read(paste0(Input,list_EC_result_250m[j]))
    data_rescaled <- data_to_rescale %>%
      dplyr::select(grid_id, Total) %>%
      inner_join(inside_grid, by = "grid_id") %>% 
      st_drop_geometry() %>% 
      group_by(grid_id_3km) %>%
      summarise(Total_3k = mean(Total, na.rm = TRUE)) %>%
      mutate_all(~ifelse(is.nan(.), NA, .)) %>% 
      drop_na(grid_id_3km) %>%
      full_join(grid_polygons_sf_3km, by = "grid_id_3km") %>%
      rename(grid_id = grid_id_3km)
    
    
    data_rescaled_classified <- QClassify(data_rescaled,data_rescaled$Total_3k, "Total")
    data_rescaled_classified$EC <- list_EC_result_250m[j]
    data_rescaled_classified <- data_rescaled_classified[,c("grid_id","Total","EC", "geometry")]
    data_EC_rescaled <- rbind(data_EC_rescaled, data_rescaled_classified)
  }
}

```

## 4. Computation of the final EV score 

```{r score_combine}

### Load shapefile with the results from each EC ###

data_EC <- data.frame()
for(i in 1:length(list_EC_result_3km)){
  data_current <- st_read(paste0(Input,list_EC_result_3km[i]))
  data_current <- data_current[,c("grid_id","Total","geometry","conf_score","nb_sample")]
  data_current$EC <- list_EC_result_3km[i]
  data_EC <- rbind(data_EC,data_current) 
}

if(exists('data_EC_rescaled') && is.data.frame(get('data_EC_rescaled'))){
  
  data_geometry <- st_as_sf(data_EC_rescaled)
  st_crs(data_geometry) <- st_crs(data_EC)
  data_EC <- rbind(data_EC,data_geometry)
}

BBT_EC_EV <- data_EC %>%
  group_by(grid_id, geometry) %>% 
  summarise(Total_EV = mean(Total, na.rm = TRUE),
            conf_score_EV = mean(conf_score, na.rm = TRUE),
            total_sample = sum(nb_sample,na.rm = TRUE),
            EC_used = sum(!is.na(conf_score)))

# Export the results to a shapefile
st_write(BBT_EC_EV, file.path(Output,"BBT_final_EV_layers.shp"), append = FALSE)


```

## 5. Map of the EV score

**Plot EV score**

```{r plot_final_score, message=FALSE, warning=FALSE, results='hide'}
if (continous == TRUE){
  print(ggplot()+
   geom_sf(data = BBT_EC_EV$geometry,lwd = 0) +
   geom_sf(data = BBT_EC_EV, aes(fill=BBT_EC_EV$Total_EV), colour = "grey90", lwd = 0) + 
   scale_fill_viridis_c(name = "Ecological Valuation score", na.value = "grey94", limits = c(0,5)) +
   geom_sf(data = world_data$geometry) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
         panel.background = element_blank(),
         axis.text.x=element_text(angle=300,hjust=0)))
ggsave(filename=file.path(Output, "Total_Valutation_Score.png"),
    width = NA,
    height = NA,
    units = "cm",
    dpi=1600)
} else{
  # Define the categories and colors
  categories <- c("0 - 1", "1 - 2", "2 - 3", "3 - 4", "4 - 5")
  mycolors <- viridisLite::viridis(5) 
  # Create categories with explicit levels
  BBT_EC_EV$category_graph <- cut(
      BBT_EC_EV$Total_EV,
      breaks = c(0, 1, 2, 3, 4, 5),
      include.lowest = TRUE,
      labels = categories,
      # Force creation of all levels
      right = FALSE,
      # Ensure that all levels are present
      levels = categories
  )
  print(ggplot()+
   geom_sf(data = BBT_EC_EV$geometry,lwd = 0) +
   geom_blank() +
   geom_sf(data = BBT_EC_EV, aes(fill=category_graph), colour = "grey90", lwd = 0, show.legend = TRUE) + 
     # show.legend allow to have colors in legend even for categories without data points
   scale_fill_manual(values = mycolors, name = "Ecological Valuation score", 
                     limits = categories, na.value = "grey94", 
                     labels = categories, drop = FALSE) +
   geom_sf(data = world_data$geometry) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
         panel.background = element_blank(),
         axis.text.x=element_text(angle=300,hjust=0))
   )
ggsave(filename=file.path(Output, "Total_Valutation_Score.png"),
    width = NA,
    height = NA,
    units = "cm",
    dpi=1600)
}

```

**Plot EV confidence score**

``` {r plot_conf_score}

if (continous == TRUE){
  print(ggplot()+
   geom_sf(data = BBT_EC_EV$geometry,lwd = 0) +
   geom_sf(data = BBT_EC_EV, aes(fill=conf_score_EV), colour = "grey90", lwd = 0) + 
   scale_fill_viridis_c(name = "Confidence Score", na.value = "grey94", limits = c(0,5)) +
   geom_sf(data = world_data$geometry) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
         panel.background = element_blank(),
         axis.text.x=element_text(angle=300,hjust=0)))
ggsave(filename=file.path(Output, "Total_confidence_score.png"),
    width = NA,
    height = NA,
    units = "cm",
    dpi=1600)
} else{
  # Define the categories and colors
  categories <- c("0 - 1", "1 - 2", "2 - 3", "3 - 4", "4 - 5")
  mycolors <- viridisLite::viridis(5) 
  # Create categories with explicit levels
  BBT_EC_EV$category_graph <- cut(
      BBT_EC_EV$conf_score_EV,
      breaks = c(0, 1, 2, 3, 4, 5),
      include.lowest = TRUE,
      labels = categories,
      # Force creation of all levels
      right = FALSE,
      # Ensure that all levels are present
      levels = categories
  )
  print(ggplot()+
   geom_sf(data = BBT_EC_EV$geometry,lwd = 0) +
   geom_blank() +
   geom_sf(data = BBT_EC_EV, aes(fill=category_graph), colour = "grey90", lwd = 0, show.legend = TRUE) + 
     # show legend allow to have colors in legend even for categories without data points
   scale_fill_manual(values = mycolors, name = "Confidence Score", 
                     limits = categories, na.value = "grey94", 
                     labels = categories, drop = FALSE) +
   geom_sf(data = world_data$geometry) +
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
         panel.background = element_blank(),
         axis.text.x=element_text(angle=300,hjust=0))
   )
ggsave(filename=file.path(Output, "Total_confidence_score.png"),
    width = NA,
    height = NA,
    units = "cm",
    dpi=1600)
}

```
